---
title: "HW_1_bhigginbotham"
author: "brian higginbotham"
date: "2024-07-02"
urlcolor: blue
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, error = FALSE)
```

## 11.22

```{r treadmill data}
treadmill <- read.table(file="data/fitness.txt", header=TRUE)
```

### a) plot the data

```{r plot treadmill}
plot(treadmill$treadmill, treadmill$tenk)
```

### b) Fit regression model to the data

```{r fit regression}
slr_tread <- lm(tenk ~ treadmill, data=treadmill)
plot(treadmill$treadmill, treadmill$tenk)
abline(slr_tread, col=2)
```
The linear model does seem appropriate - there is an overall downward trend in the data that is reflected in the model.

### c) estimate the linear regression model

```{r model summary}
summary(slr_tread)
```
We see from the summary output that the estimates for the linear model are `b_hat_0 (Intercept) = 58.8158` and `b_hat_1 (slope) = -1.8673`. Therefore the estimate for the linear regression model would be

`y_hat = 58.8158 - 1.8673x_i` for any given `i`.

## 11.23

### a) estimate population variance for errors (sigma^2_e)

`Residual Standard Error (RSE) = 2.102`  
`RSE = sqrt(sigma^2_e)`  
`RSE^2 = sigma^2_e`
```{r error variance}
RSE <- 2.102
RSE**2
```
Estimate for error variance

### b) estimate standard error for b_hat_1

`0.3462` - from summary output: Std. Error for 'treadmill'

### c) 95% Confidence Interval for b_hat_1
```{r b_hat_1 CI}
confint(slr_tread, 'treadmill', level=0.95)
```
the plausible values of b_1, given a significance level of 0.05 would be
`-2.59 < b_1 < -1.140`

### d) hypothesis test that there is a linear relationship between treadmill time and 10K time
`Pr(>|t|)` for 'treadmill' from summary output = `3.99e-05` - near zero percent chance of the null hypothesis (b_hat_1 = 0) occuring for the t value `-5.393` in the student t-distribution.  
**Reject H_0** - conclude there is a linear relationship between treadmill time and 10K time